---
title: "COVID prediction - model"
author: "Robert On, Hal Varian"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    code_folding: hide
---

# Model

This notebook explores different models for predicting US county-level COVID cases and deaths. This includes transformations of the response variable (log, square root, logit, etc) as well as trading off the inclusion of different predictors and the effects of missingness on our sample of counties. We leverage variable selection via Spike-and-Slab to provide sparsity in our models to avoid overfitting.

### Missing values

As suggested in the tables above, each of these columns are missing some predictors since there isn't full coverage for all counties. There is a trade off in having more predictors vs more counties. The graph below helps us manage this trade off.

```{r}
missing_seq <- seq(0.01, 0.99, by = 0.01)
lapply(missing_seq, FUN = function(thresh) {
  d <- readd(county_joined_all) %>%
    select_if(~ mean(is.na(.x)) < thresh) %>%
    select(-starts_with("jhu-csse_")) %>%
    drop_na() %>%  dim()
  list(counties = d[1], predictors = d[2], thresh = thresh)
}) %>% bind_rows() %>%
  ggplot(aes(predictors, counties, label=thresh)) + geom_line() -> p
ggplotly(p)
```

From a quick eye-ball it appears optimal to use 2170 counties with 101 predictors which suggests a threshold limiting us to use predictors with less than 31% missing values (`county_joined_rich`). We can also experiment with a model with 3118 counties and 46 predictors with a 3% threshold (`county_joined_sparse`). A third model is one that has a full set of predictors (139) but only covers 295 counties (`county_joined_full`) that allows for 92% missing values.`

### Outcomes

The folks at the Delphi project have convenient provided COVID case and death counts from JHU in their data set.

```{r}
thresh_data <- function(thresh) {
  readd(county_joined_all) %>%
    select_if(~ mean(is.na(.x)) <= thresh) %>% drop_na() %>%
    unite(name, name_county, name_state, sep = ", ") %>%
    select(-state_fips, -fips5) %>%
    column_to_rownames("name") %>%
    rename_at(vars(ends_with(" raw value")), ~ str_remove(.x, " raw value")) %>%
    rename_at(vars(ends_with("_percent_change_from_baseline")),
              ~ str_remove(.x, "_percent_change_from_baseline"))
}

county_joined_full <- thresh_data(0.92)
county_joined_rich <- thresh_data(0.31)
county_joined_sparse <- thresh_data(0.03)
```

With our three data sets (sparse, rich and full) and two outcomes (proportion of confirmed incidence and deaths), we can go ahead and make predictions using our set of predictors. We start with a regularized regression with the [Spike-and-Slab](https://www.rdocumentation.org/packages/BoomSpikeSlab/versions/1.2.3/topics/lm.spike) package which provides a Bayesian approach towards variable selection similar to the LASSO. We also try square root and logit transformations of the response variables.


```{r}
bind_rows(
  county_joined_full %>% select(starts_with("jhu-csse_"), chr_Population) %>%
    mutate(predictors = ncol(county_joined_full), counties = n()) %>%
    summarise_all(mean, na.rm = TRUE) %>% mutate(dataset = "full"),
  county_joined_rich %>% select(starts_with("jhu-csse_"), chr_Population) %>%
    mutate(predictors = ncol(county_joined_rich), counties = n()) %>%
    summarise_all(mean, na.rm = TRUE) %>% mutate(dataset = "rich"),
  county_joined_sparse %>% select(starts_with("jhu-csse_"), chr_Population) %>%
    mutate(predictors = ncol(county_joined_sparse), counties = n()) %>%
    summarise_all(mean, na.rm = TRUE) %>% mutate(dataset = "sparse")
) %>% select(dataset, predictors, counties, chr_Population, everything()) %>%
  rename_at(vars(starts_with("jhu-csse_")), ~ str_remove(.x, "jhu-csse_"))
```

We notice in general that COVID cases and deaths are lower in the `sparse` data compared to the `full` data. It's likely that the sparse set contains smaller counties that had more missing predictors. The incidence and death proportions are closer in value between the two sets and looks to be a more resilient outcome to use.

### All variables

Varied rows based on NAs.

```{r message=FALSE, warning=FALSE, results="hide"}
iters <- 10000

deaths_full <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(`jhu-csse_deaths_incidence_prop` ~ ., data = ., niter=iters)
deaths_rich <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(`jhu-csse_deaths_incidence_prop` ~ ., data = ., niter=iters)
deaths_sparse <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(`jhu-csse_deaths_incidence_prop` ~ ., data = ., niter=iters)

cases_full <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)
cases_rich <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)
cases_sparse <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)

deaths_full_sqrt <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(sqrt(`jhu-csse_deaths_incidence_prop`) ~ ., data = ., niter=iters)
deaths_rich_sqrt <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(sqrt(`jhu-csse_deaths_incidence_prop`) ~ ., data = ., niter=iters)
deaths_sparse_sqrt <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(sqrt(`jhu-csse_deaths_incidence_prop`) ~ ., data = ., niter=iters)

cases_full_sqrt <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(sqrt(`jhu-csse_confirmed_incidence_prop`) ~ ., data = ., niter=iters)
cases_rich_sqrt <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(sqrt(`jhu-csse_confirmed_incidence_prop`) ~ ., data = ., niter=iters)
cases_sparse_sqrt <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(sqrt(`jhu-csse_confirmed_incidence_prop`) ~ ., data = ., niter=iters)

deaths_full_log <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(log(`jhu-csse_deaths_incidence_prop` + 1) ~ ., data = ., niter=iters)
deaths_rich_log <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(log(`jhu-csse_deaths_incidence_prop` + 1) ~ ., data = ., niter=iters)
deaths_sparse_log <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  lm.spike(log(`jhu-csse_deaths_incidence_prop` + 1) ~ ., data = ., niter=iters)

cases_full_log <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(log(`jhu-csse_confirmed_incidence_prop` + 1) ~ ., data = ., niter=iters)
cases_rich_log <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(log(`jhu-csse_confirmed_incidence_prop` + 1) ~ ., data = ., niter=iters)
cases_sparse_log <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(log(`jhu-csse_confirmed_incidence_prop` + 1) ~ ., data = ., niter=iters)

deaths_full_logit <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  logit.spike(`jhu-csse_deaths_incidence_prop` ~ ., data = ., niter=iters)
deaths_rich_logit <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  logit.spike(`jhu-csse_deaths_incidence_prop` ~ ., data = ., niter=iters)
deaths_sparse_logit <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_deaths_incidence_prop`) %>%
  logit.spike(`jhu-csse_deaths_incidence_prop` ~ ., data = ., niter=iters)

cases_full_logit <- county_joined_full %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  logit.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)
cases_rich_logit <- county_joined_rich %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  logit.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)
cases_sparse_logit <- county_joined_sparse %>%
  select(-starts_with("jhu-csse_"), `jhu-csse_confirmed_incidence_prop`) %>%
  logit.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)
```

```{r}
mdls <- list(cases_full, deaths_full, cases_rich, deaths_rich,
             cases_sparse, deaths_sparse,
             cases_full_sqrt, deaths_full_sqrt, cases_rich_sqrt,
             deaths_rich_sqrt, cases_sparse_sqrt, deaths_sparse_sqrt,
             cases_full_log, deaths_full_log, cases_rich_log,
             deaths_rich_log, cases_sparse_log, deaths_sparse_log,
             cases_full_logit, deaths_full_logit, cases_rich_logit,
             deaths_rich_logit, cases_sparse_logit, deaths_sparse_logit)
mdl_names <- c("cases_full", "deaths_full", "cases_rich", "deaths_rich",
               "cases_sparse", "deaths_sparse",
               "cases_full_sqrt", "deaths_full_sqrt", "cases_rich_sqrt", 
               "deaths_rich_sqrt",
               "cases_sparse_sqrt", "deaths_sparse_sqrt",
               "cases_full_log", "deaths_full_log", "cases_rich_log", 
               "deaths_rich_log",
               "cases_sparse_log", "deaths_sparse_log",
               "cases_full_logit", "deaths_full_logit", "cases_rich_logit",
               "deaths_rich_logit",
               "cases_sparse_logit", "deaths_sparse_logit")
rsq_df <-bind_rows(
  tibble(
    model = mdl_names[1:18],
    predictors = mdls[1:18] %>% map_int(~ ncol(.x$training.data)),
    counties = mdls[1:18] %>% map_int(~ nrow(.x$training.data)),
    rsq = mdls[1:18] %>% map_dbl(~ summary(.x)$rsquare[["Mean"]])),
  tibble(
    model = mdl_names[19:24],
    predictors = mdls[19:24] %>% map_int(~ ncol(.x$training.data)),
    counties = mdls[19:24] %>% map_int(~ nrow(.x$training.data)),
    rsq = mdls[19:24] %>% map_dbl(~ summary(.x)[["deviance.r2"]])))

rsq_df
```

```{r}
tmp <- sapply(1:length(mdls), FUN = function(x) {
  mdl <- mdls[[x]]
  rsq <- rsq_df$rsq[x]
  plot(mdl, "inclusion", inclusion.thresh = 0.25,
       main = paste(mdl_names[x], paste0(nrow(mdl$training.data), " counties"),
                    paste0("rsq: ", round(rsq, 2)), sep = " - ")) })
```

### Dataset predictive value

What is the predictive value for the mobility, community health rankings, and COVIDcast data? Only the full model has variables from all data sets.

```{r message=FALSE, warning=FALSE, results="hide"}
cases_full_mob <- county_joined_full %>%
  select(starts_with("mob_"), -starts_with("jhu-csse_"),
         `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)

cases_full_chr <- county_joined_full %>%
  select(starts_with("chr_"), -starts_with("jhu-csse_"),
         `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)

cases_full_cmu <- county_joined_full %>%
  select(starts_with("cmu_"), -starts_with("jhu-csse_"),
         `jhu-csse_confirmed_incidence_prop`) %>%
  lm.spike(`jhu-csse_confirmed_incidence_prop` ~ ., data = ., niter=iters)
```

```{r}
mdls <- list(cases_full_mob, cases_full_chr,
             cases_full_cmu)
mdl_names <- c("cases_full_mob", "cases_full_chr",
               "cases_full_cmu")
rsq_df <- tibble(
  model = mdl_names,
  predictors = mdls %>% map_int(~ ncol(.x$training.data)),
  counties = mdls %>% map_int(~ nrow(.x$training.data)),
  rsq = mdls %>% map_dbl(~ summary(.x)$rsquare[["Mean"]]))
rsq_df
```

```{r}
tmp <- sapply(1:length(mdls), FUN = function(x) {
  mdl <- mdls[[x]]
  rsq <- rsq_df$rsq[x]
  plot(mdl, "inclusion", inclusion.thresh = 0.25,
       main = paste(mdl_names[x], paste0(nrow(mdl$training.data), " counties"),
                    paste0("rsq: ", round(rsq, 2)), sep = " - ")) })
```

## Reproducibility

```{r}
## datetime
Sys.time() 

## repository
git2r::repository()

## session info
sessionInfo()
```
